---
title: "CUNY621_FinalProject"
author: "Zachary Herold"
date: "April 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

#### Loading the data

```{r}
white <- data.frame(read.csv("https://raw.githubusercontent.com/jjohn81/DATA621_Final_Project/master/winequality-white.csv", sep = ";"))
```


The following is the breakdown by quality rating. 

```{r}
table(factor(white$quality))
```

Consider consolidating the wines into five ordinal categories. 

inferior = 3,4
subpar = 5
average = 6
abovepar = 7
superior = 8,9


## Ordinal Logistic Regression
```{r wines}
m1 <- polr(factor(quality) ~., data=white, Hess=TRUE)
summary(m1)
```

This regression output coefficient table includes the value of each coefficient, standard errors, and t value, which is simply the ratio of the coefficient to its standard error. 

One notes the estimates for the six intercepts (for each one-point step in quality improvement, from 3 to 9), which are sometimes called cutpoints. The intercepts indicate where the latent variable is cut to make the seven groups that we observe in our data. 

There is no significance test in the 'polr' function by default, but we can generate p-values by comparing the t-value against the standard normal distribution, like a z test. Of course this is only true with infinite degrees of freedom, but is reasonably approximated by large samples (in this dataset, nrow(white) = 4898). 

```{r}
ctable <- coef(summary(m1))
## calculate and store p values
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
ctable <- cbind(ctable, "p value" = p)
ctable
```

Below, we computer coefficients, which are called proportional odds ratios (OR) and we would interpret these pretty much as we would odds ratios from a binary logistic regression.

```{r}
ci <- confint(m1)
exp(cbind(OR = coef(m1), ci))
```

